---
title: ""
author: "Zack Loken"
date: "February 6, 2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Zack Loken Lab Exercise 2 - Data Analysis in R


### 1. What is the second argument in the lm() function?
      
##### The second argument in the lm() function is data - which allows you to manually enter values for the model variables. This was found by running the command ?lm() and then reviewing the arguments listed in the R documentation for Fitting Linear Models.
    
```{r eval=FALSE}
?lm()
```


### 2. What is the second argument in the plot.gam() function found in the mgcv library? 
    
##### The second argument in the plot.gam() function is residuals; if true then partial residuals are added to plots. This was found by installing the mgcv package and then typing ?plot.gam(). We can then review arguments listed in the R documentation for Default GAM plotting.

```{r echo=FALSE}
chooseCRANmirror(ind=81)
```

```{r eval=FALSE}
install.packages("mgcv")
library(mgcv)
?plot.gam()
```


### 3. What is the length of the vector created from a sequence of 3 to 27 by intervals of 0.13?
    
##### The vector length of this output is 185. 

```{r}
v <- seq(3, 27, by=0.13)
length(v)
```


### 4. What code subsets the first row and third column of the second matrix for the array 'array(seq(1:20)), dim=c(2,5,2))' and what number occupies this position?

##### The code that subsets the first row and third column of the second matrix for this array is 'a[1, 3, 2]' and the number that occupies this position is 15.

```{r}
a <- array(seq(1:20), dim=c(2,5,2))
a[1, 3, 2]
```


### 5. Create a list object from 3 or more built in datasets.

```{r}
data(package = .packages(all.available = TRUE))
data("rivers")
data("USArrests")
data("precip")
data("rock")
ListObject <- list(rivers, USArrests, precip, rock)
str(ListObject)
```


### 6. What is the mean length of major North American Rivers ('rivers' built-in dataset)

##### The mean length of major North American Rivers is 591.2 units

```{r}
data("rivers")
summary(rivers)
```


### 7. Write a built-in dataset to a CSV file (show code) and open it in Excel. 

```{r}
data("PlantGrowth")
write.csv(PlantGrowth, 
          "D:/Zack/School/Classes/Spring 22/OCS 7131 - Data Analysis in R/Lab/Lab 1/PlantGrowth.csv", 
          row.names=TRUE)
```


### 8. Tabulate the type and concentration factors in the CO2 dataset in one table. Is this experiment balanced?

##### See code below. This experiment is balanced because there are an equal number of observations at each concentration level for Quebec and Mississippi (values in the 'Type' column). 

```{r}
data("CO2")
table(CO2$Type, CO2$conc)
```


### 9. Write a single function to subset both spray A and records of counts > 17 from 'InsectSprays'. How many records are there? 

##### There are 3 records with spray type A and count greater than 17. 

```{r}
data("InsectSprays")
subset(InsectSprays, spray=='A' & count > 17, select = c(spray, count))
```


### 10. Create a box plot of the spray counts in 'InsectSprays'.

```{r}
data("InsectSprays")
boxplot(count ~ spray, data = InsectSprays,
        xlab = "Spray Type",
        ylab = "Spray Count",
        col = "lightblue")
```