---
title: ""
author: "Zack Loken"
date: "2/21/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

<center> <h1>__Zack Loken - OCS 7131 Exercise 4__</h1> </center>
<br>

```{r message=FALSE}
library(tidyverse)
```
<br>

#### 1. Use the `pivot_longer()` function to turn the following dataframe into one with a single column for year. 

```{r}
Species <- c(rep("A",3),rep("B",3),rep("C",3))
set.seed(16)
Year2017 <- rpois(n = 9, lambda = 1)
Year2018 <- rpois(n = 9, lambda = 3)
Year2019 <- rpois(n = 9, lambda = 5)
Year2020 <- rpois(n = 9, lambda = 7)
wide.df <- as.data.frame(cbind(Species, Year2017, Year2018, Year2019, Year2020))

narrow.df1 <- pivot_longer(data = wide.df, 
                          cols = c(2, 3, 4, 5), 
                          names_to = "Year", 
                          names_prefix = "Year", 
                          values_to = "Count")
```
<br>

#### 2. Use the `str_sub()` function to remove the term "Year" from data values in the "Year" column. 
      
####### Using the `pivot_longer()` function, you can define the parameter `names_prefix` = "Year" to remove matching text from the start of each variable name for the columns being pivoted into narrow format. This accomplishes the same thing as using the `str_sub()` function. See below for how to do it with `str_sub()`.

```{r}
# Transform wide dataframe into narrow format
narrow.df2 <- pivot_longer(data = wide.df, 
                          cols = c(2, 3, 4, 5), 
                          names_to = "Year", 
                          values_to = "Count")

# Use the str_sub() function to create a new column with only the numeric year
narrow.df2$substring_Year<- str_sub(narrow.df2$Year, 5, 8)
```
<br>

#### 3. Use the `summarise()` function to report the mean number of each species for each year. You'll want to consider the `group_by()` function as well.  

```{r}
# Start by converting your Year and Count columns to numeric type
narrow.df1[, c(2:3)] <- sapply(narrow.df1[, c(2:3)], as.numeric)

# Group by species and year, then summarize the mean of each species for each year.  
sppCountAvg <- narrow.df1 %>% 
                group_by(Species, Year) %>%
                summarise(AvgCount = mean(Count), n = n())
```
<br>

#### 4. Use `ggplot()` to plot species against year. Add some information, like color or shape, to indicate specific species. 

```{r}
ggplot(sppCountAvg, aes(Year, AvgCount)) +
  geom_col(aes(fill = Species), position = "dodge") +
  labs(x = "Observation Year",
       y = "Average Species Count",
       caption = "A column chart of average species count for 3 species
         across 4 years. Points are color coded by species type.",
       color = "Species",
       title = "Average Species Count Between 2017 and 2020 for 3 Species") +
   theme_classic()
```