---
title: ""
author: "Zack Loken"
date: "3/7/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.align = "CENTER", fig.width = 14)
```

<center> <h1>__Zack Loken - OCS 7131 Exercise 5__</h1> </center>
<br>

##### Figure 1. Use `ggplot2` to create a column chart showing total surface water area annually by covertype class from 1984 to 2020 for a portion of Missouri's Lower Grand River floodplain. 
```{r}
library(tidyverse)
library(RColorBrewer)
library(data.table)

# Loading the data
LGRW_SWA <- read.csv("D:/Zack/Google Earth Engine/GEE for Systems Ecology - Spring 2021/Missouri Habitat and Water Classification/SurfaceWaterArea_Annual/LGRW_WetHectares_byYear.csv")

# Quick cleanup -- rename the column headers
setnames(LGRW_SWA, skip_absent = TRUE,
         old = c('CoverType', 'Owner', 'Hectares', 'X1984_SWA', 'X1987_SWA', 'X1989_SWA', 'X1990_SWA', 'X1991_SWA', 
                 'X1992_SWA', 'X1994_SWA', 'X1996_SWA', 'X1997_SWA', 'X1999_SWA', 'X2001_SWA', 'X2004_SWA', 'X2005_SWA',
                 'X2006_SWA', 'X2010_SWA', 'X2011_SWA', 'X2014_SWA', 'X2015_SWA', 'X2016_SWA', 'X2018_SWA', 'X2019_SWA',
                 'X2020_SWA'),
         new = c('Class', 'Owner', 'Hectares', '1984', '1987', '1989', '1990', '1991', '1992', '1994', '1996', '1997', 
                 '1999', '2001', '2004', '2005', '2006', '2010', '2011', '2014', '2015', '2016', '2018', '2019', '2020'))

# Subset to only show covertype classes with high annual surface water area.  
LGRW_SWA_Sub <- subset(LGRW_SWA, Class == c('Rotational Wet', 'Grand River', 'Wetland', 'Wet Meadow',
                      'Forest - Riprarian', 'Forest - Hardmast', 'Cultivated', 'Water Bodies'), 
                       select = c(1:25))

# Convert wide data frame to narrow format with one column containing year. 
narrow.LGRW_SWA_Sub <- pivot_longer(data = LGRW_SWA_Sub, 
                          cols = c(4:25), 
                          names_to = "Year", 
                          values_to = "SWA_Hectares")

# Remove 'Hectares' column from narrow.LGRW_SWA dataframe
narrow.LGRW_SWA_Sub2 <- subset(narrow.LGRW_SWA_Sub, select = -Hectares)

# Group by class and year, then summarize the total surface water area of each class for each year.  
LGRW_SWA_Sum <- narrow.LGRW_SWA_Sub2 %>% 
                    group_by(Class, Year) %>%
                    summarise(SWA_sum = sum(SWA_Hectares), n = n())

# Change all 'Forest - Riprarian' values to 'Forest - Riparian'
LGRW_SWA_Sum$Class[45:66] <- "Forest - Riparian"

# Plot total surface water area of each class by year. 
ggplot(LGRW_SWA_Sum, aes(Year, SWA_sum)) +
  geom_col(aes(fill = Class), position = "dodge") +
  labs(x = "Year",
       y = "Total Surface Water Area (Hectares)",
       caption = "Surface water area (hectares) by landcover 
       type in Missouri's Lower Grand River floodplain.",
       color = "Class",
       title = "Annual Surface Water Area in Missouri's Lower Grand River floodplain") +
  scale_fill_manual(values = brewer.pal(8, "Dark2")) + theme_classic(base_size = 14)
```
<br>

##### Figure 2. Use `ggplot2` to create a pie chart showing total surface water area (%) of each covertype for all years combined in a portion of Missouri's Lower Grand River floodplain.
```{r}
library(tidyverse)
library(RColorBrewer)
library(data.table)
library(formattable)

# Loading the data
LGRW_SWA <- read.csv("D:/Zack/Google Earth Engine/GEE for Systems Ecology - Spring 2021/Missouri Habitat and Water Classification/SurfaceWaterArea_Annual/LGRW_WetHectares_byYear.csv")

# Quick cleanup -- rename the column headers
setnames(LGRW_SWA, skip_absent = TRUE,
         old = c('CoverType', 'Owner', 'Hectares', 'X1984_SWA', 'X1987_SWA', 'X1989_SWA', 'X1990_SWA', 'X1991_SWA', 
                 'X1992_SWA', 'X1994_SWA', 'X1996_SWA', 'X1997_SWA', 'X1999_SWA', 'X2001_SWA', 'X2004_SWA', 'X2005_SWA',
                 'X2006_SWA', 'X2010_SWA', 'X2011_SWA', 'X2014_SWA', 'X2015_SWA', 'X2016_SWA', 'X2018_SWA', 'X2019_SWA',
                 'X2020_SWA'),
         new = c('Class', 'Owner', 'Hectares', '1984', '1987', '1989', '1990', '1991', '1992', '1994', '1996', '1997', 
                 '1999', '2001', '2004', '2005', '2006', '2010', '2011', '2014', '2015', '2016', '2018', '2019', '2020'))

# Subset to only show covertype classes with high annual surface water area.  
LGRW_SWA_Sub <- subset(LGRW_SWA, Class == c('Rotational Wet', 'Grand River', 'Wetland', 'Wet Meadow',
                      'Forest - Riprarian', 'Forest - Hardmast', 'Cultivated', 'Water Bodies'), 
                       select = c(1:25))

# Convert wide data frame to narrow format with one column containing year. 
narrow.LGRW_SWA_Sub <- pivot_longer(data = LGRW_SWA_Sub, 
                          cols = c(4:25), 
                          names_to = "Year", 
                          values_to = "SWA_Hectares")

# Remove 'Hectares' column from narrow.LGRW_SWA dataframe
narrow.LGRW_SWA_Sub2 <- subset(narrow.LGRW_SWA_Sub, select = -Hectares)

# Group by class, then summarize the total surface water area of each class for each year.  
LGRW_SWA_Sum <- narrow.LGRW_SWA_Sub2 %>% 
                    group_by(Class) %>%
                    summarise(SWA_sum = sum(SWA_Hectares), n = n())

# Change all 'Forest - Riprarian' values to 'Forest - Riparian'
LGRW_SWA_Sum$Class[3] <- "Forest - Riparian"

# Convert total surface water area to a percent per covertype. 
LGRW_SWA_Per <- LGRW_SWA_Sum %>%
  add_column(Percent_Cover = ((LGRW_SWA_Sum$SWA_sum[1 & 2 & 3 & 4 & 5 & 6 & 7 & 8]/sum(LGRW_SWA_Sum$SWA_sum[1:8])*100)))

# Create a bar plot showing total surface water area (%) of each class for all years combined.
bp <- ggplot(LGRW_SWA_Per, aes(x = "", y = Percent_Cover, fill = Class)) + 
  geom_bar(width = 1, stat = "identity") + 
  labs(x = NULL, 
       y = NULL,
       title = "Percent Surface Water Area by Covertype") + 
  scale_fill_manual(values = brewer.pal(8, "Dark2")) + 
  theme_minimal(base_size = 14)

# Create a pie chart showing total surface water area (%) of each class for all years combined. 
pie <- bp + coord_polar("y", start = 0)
pie
```
