---
title: ""
author: "Zack Loken"
date: "3/15/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.align = "CENTER", fig.width = 14)
```

<center> <h1>__Zack Loken - OCS 7131 Exercise 6__</h1> </center>
<br>

#### __1. When pooling across all lakes and years, does secchi depth increase, decrease, or not change throughout the summer months?__


```{r}
# Load the library that contains the lake quality data
library("adklakedata")

# Load the secchi depth data to an object named 'secchi'
secchi <- adk_data('secchi')

# (A) Consider a linear model where months are a continuous predictor.

## Simple linear regression is a suitable model; it tests for the 
## effect of a continuous predictor on the response variable.

### month - continuous - predictor variable (x);
### secchi - continuous - response variable (y)

# Isolate columns into separate objects and define their data types. 
month = as.numeric(secchi$month)
secchi = as.numeric(secchi$secchi)

# Define the linear model.
A1 <- lm(secchi ~ month)
A1

# Evaluate the residuals and model fit, and any reported significance. 
summary.lm(A1)

# Plot the residuals to a histogram to see if they are close to being 
# centered over 0 w/ roughly symmetric shape. 
resA1 <- resid(A1)
hist(resA1, breaks = 10, las = 1, col = 1, border = "white", xlab = "Residual", main = '')

# Plot the residuals as a function of the fitted data. 
# Look for residuals spread evenly, and w/ symmetry, above and below 0
fitA1 <- fitted(A1)
plot(resA1 ~ fitA1, pch = 19, las = 1); abline(0, 0, col = "red")
```
<br>

##### (A). The above linear model (simple linear regression) considered month to be a continuous predictor. When pooling across all lakes and years, mean secchi depth decreases throughout the summer months. The `month` estimate of -0.0759 tells us that for every 1 unit increase in month, mean secchi depth decreases by 0.0759. The intercept was 5.068. We can conclude that water clarity does not significantly change throughout the summer because of the negative R-squared value (R-squared = -0.0003269) and a p-value > 0.05 (p = 0.406). 
<br>

```{r}
# Load the secchi depth data to an object named 'secchi'
secchi <- adk_data('secchi')

# (B) Consider a linear model where months are a factor (categorical predictor). 

## ANOVA is a suitable model; it tests for the effect of a categorical
## predictor with 3 or more groups on the response variable. 

### month - categorical - predictor w/ 4 factor levels (x); 
### secchi - continuous- response variable (y)

# Isolate columns into separate objects and define their data types. 
month = as.factor(secchi$month)
secchi = as.numeric(secchi$secchi)

# Define the linear model.
B1 <- aov(secchi ~ month)
B1

# Evaluate the residuals and model fit, and any reported significance. 
summary(B1)

# Plot the residuals to a histogram to see if they are close to being 
# centered over 0 w/ roughly symmetric shape. 
resB1 <- resid(B1)
hist(resB1, breaks = 10, las = 1, col = 1, border = "white", xlab = "Residual", main = '')

# Plot the residuals as a function of the fitted data. 
# Look for residuals spread evenly, and w/ symmetry, above and below 0
fitB1 <- fitted(B1)
plot(resB1 ~ fitB1, pch = 19, las = 1); abline(0, 0, col = "red")

```
<br>

##### (B). The above linear model (ANOVA) considered month to be a categorical predictor with 4 factor levels. When pooling across all lakes and years, mean secchi depth decreases from June to July, then increases each consecutive month from July to September. The F-statistic for the ANOVA is 2.04 on 3 and 941 degrees of freedom with a p-value > 0.05 (p = 0.107). Because the results of the ANOVA were insignificant, the null hypothesis can be accepted, and therefore do not need to run a post-hoc multiple comaprisons test. Based on the histrogram, it is evident that the distribution is Poisson (mode = 0, data pushed up against 0 w/ tail extending to the right) rather than normal (i.e., symmetric). 
<br>

#### __2. When pooling across all lakes and all months, does secchi depth increase, decrease, or not change throughout the years in the data (1994 - 2012)?__

```{r}
# Load the secchi depth data to an object named 'secchi'
secchi <- adk_data('secchi')

# Consider a linear model where years are a continuous predictor.

## Simple linear regression is a suitable model; it tests for the 
## effect of a continuous predictor on the response variable.

### year - continuous - predictor variable (x);
### secchi - continuous - response variable (y)

# Isolate columns into separate objects and define their data types. 
year = as.numeric(secchi$year)
secchi = as.numeric(secchi$secchi)

# Define the linear model.
A2 <- lm(secchi ~ year)
A2

# Evaluate the residuals and model fit, and any reported significance. 
summary.lm(A2)

# Plot the residuals to a histogram to see if they are close to being 
# centered over 0 w/ roughly symmetric shape. 
resA2 <- resid(A2)
hist(resA2, breaks = 10, las = 1, col = 1, border = "white", xlab = "Residual", main = '')

# Plot the residuals as a function of the fitted data. 
# Look for residuals spread evenly, and w/ symmetry, above and below 0
fitA2 <- fitted(A2)
plot(resA2 ~ fitA2, pch = 19, las = 1); abline(0, 0, col = "red")
```
<br>

##### The above linear model (simple linear regression) considered year to be a continuous predictor. When pooling across all lakes and years, mean secchi depth decreases across years. The `year` estimate of -0.0406 tells us that for every 1 unit increase in year, mean secchi depth decreases by 0.0406. The intercept was 85.810. We can conclude that water clarity significantly changes accross years (i.e., we can reject the null hypothesis) because of the p-value < 0.05 (p = 0.004), but the R-squared value of 0.0077 tells us that the predictor variable year does not account for much of the mean of the response variable secchi depth. 
<br>